---
title: "Installing and setting up rPHG and rJava"
output:
  BiocStyle::html_document:
    toc: false
    fig_caption: true
    toc_float: true
    number_sections: false
vignette: >
  %\VignetteIndexEntry{Getting Start with rPHG}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\usepackage[utf8]{inputenc}
  %\usepackage{float}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  fig.path = "figure/graphics-",
  cache.path = "cache/graphics-",
  fig.align = "center",
  external = TRUE,
  echo = TRUE,
  warning = FALSE
)
```

In contrast to other R packages, installing `rPHG` involves additional steps 
that require linking the package to Java. This document will
show you how to set up `rJava` and properly load `rPHG`.

# Installation

## Prerequisite - installing `rJava`
Since the PHG API is written primarily in Java, a Java JDK will need to be
installed on your machine. Additionally, for R to communicate with Java,
the R package `rJava` will need to be installed. More info on the `rJava` 
package can be found [here](https://www.rforge.net/rJava/). In order to load 
`rJava` properly, ensure that you have:

-   A `JDK` (Java Development Kit $\geq$ `8`) installed on your system.
-   Your system environment variable `JAVA_HOME` is configured
    appropriately and points to your `JDK` of choice. This will usually
    be included in your PATH environment variable as well. Options and
    system environmental variables that are available from R can be seen
    with `Sys.getenv()` and more specifically
    `Sys.getenv("JAVA_HOME")`.

**NOTE**: If you are using a UNIX system (e.g. Ubuntu) and are experiencing 
issues, you may need to reconfigure R with Java. To perform this, open a 
terminal and enter the command:

    R CMD javareconf

You may need to have root privileges when performing this so you may
need to add `sudo` to the prior command.

If you need additional steps on how to perform these actions, detailed
information can be found using the following links, depending on your
OS:

-   [Linux](https://datawookie.netlify.com/blog/2018/02/installing-rjava-on-ubuntu/)
-   [macOS](https://zhiyzuo.github.io/installation-rJava/)
-   [Windows](https://cimentadaj.github.io/blog/2018-05-25-installing-rjava-on-windows-10/installing-rjava-on-windows-10/)


## Install from GitHub
After you have `rJava` up and running on your machine, install the latest
version from GitHub:

```{r, eval=FALSE, echo=TRUE}
# install.packages("pak")
pak::pak("maize-genetics/rPHG")
```

After the package has been installed, load `rPHG` using:

```{r, eval=FALSE, echo=TRUE}
library(rPHG)
```

Or, if you want to use a function without violating your environment you can
use `rPHG::<function>`, where `<function>` is an `rPHG` function.



# Set-up

## Setting memory
Since certain analyses can possibly use up a lot of computational
resources, memory allocation to `rPHG` can be modified. To change the amount
of memory, use the base `options()` function and modify the following parameter:

```{r, eval=FALSE, echo=TRUE}
options(java.parameters = c("-Xmx<memory>", "-Xms<memory>"))
```

| ⚠  This will need to be set before loading the `rPHG` package! |
|:--------------------------------------------------------------:|

Replace `<memory>` with a specified unit of memory. For example, if I want to
allocate a maximum of 6 GB of memory for my operations, I would use the input
`"-Xmx6g"`, where `g` stands for gigabyte (GB). More information about memory
allocation can be found [here](https://stackoverflow.com/questions/14763079/what-are-the-xms-and-xmx-parameters-when-starting-jvm).


## Starting a logging file
Since the `rPHG` package is an interface to the 
[PHG API](https://bitbucket.org/bucklerlab/practicalhaplotypegraph), we can
track internal API progress via a logging file. This file is beneficial
for debugging, tracking, and reproducing the progress of your workflow.

| ⚠  It is _highly recommended_ to set up a logging file! |
|:-------------------------------------------------------:|

To start a logging file, use the following command:

```{r, eval=FALSE, echo=TRUE}
startLogger(fullPath = NULL, fileName = NULL)
```

If the `startLogger()` parameters are set to `NULL`, the logging file
will be created in your current working directory. If you are unsure of what
your working directory is in R, use the base `getwd()` command.



# Commonly encountered issues with `rJava`

## Problems installing rJava M1/M2/M$n$ CPU architecture (macOS)

If you are running into issues with installing `rJava` using the newer
Mac chip architecture, Oracle JDK currently (as of writing this) does not 
work. Consider an alternative JDK source such as [OpenJDK](https://openjdk.org/)
or [Azul JDK](https://www.azul.com/downloads/?version=java-8-lts&package=jdk).

More detailed information about a possible workaround can be found in this
[Stack Overflow post](https://stackoverflow.com/questions/67849830/how-to-install-rjava-package-in-mac-with-m1-architecture).


## Problems with rJava if you have upgraded Java (macOS)
When using macOS, if you previously had `rJava` working through RStudio, then 
you upgraded your Java and it now longer works, try the following:

At the command line type:

```{bash, eval=FALSE, echo=TRUE}
R CMD javareconf
```

Then check for a left over symbolic link via:

```{bash, eval=FALSE, echo=TRUE}
ls -ltr /usr/local/lib/libjvm.dylib
```

If the link exists, remove it, then create it fresh via these commands:

```{bash, eval=FALSE, echo=TRUE}
rm /usr/local/lib/libjvm.dylib
sudo ln -s $(/usr/libexec/java_home)/lib/server/libjvm.dylib /usr/local/lib
```

You should now be able to enter RStudio and setup rJava.


## Problems loading the `rJava` package (Linux)
If you are using a UNIX system (e.g. Ubuntu) and are experiencing 
issues, you may need to reconfigure R with Java. To perform this, open a 
terminal and enter the command:

``` bash
R CMD javareconf
```

You may need to have root privileges when performing this so you may
need to add `sudo` to the prior command:

``` bash
sudo R CMD javareconf
```


